### 2021-11-23 draft.tex
Deploying to gh-pages from @ gottacatchenall/thesis_proposal@3f9993b383ac890764ce6e56e86cf0289a33738f 🚀

Within the last several hundred years, human activity has induced rapid
changes in Earth's atmosphere, oceans, and surface. Greenhouse gas
emissions have caused an increase the temperature of both Earth's
terrain and oceans, and both agricultural and urban development has
rapidly reshaped the Earth's land cover. These the bulk of this change
has occurred within the last several hundred years, a geological
instant, inducing a sudden shift in conditions to Earth's climate and
biosphere. As a result \emph{ecological forecasting}---TK DEFINITION
---and then using these forecasts to make decisions to mitigate the
negative consequences of this change on ecosystems, their functioning,
and the services they provide to humans has emerged as an imperative for
ecology and environmental science (Dietze 2017). However, robust
prediction of ecological processes is, to say the least, quite difficult
(Beckage \emph{et al.} 2011; Petchey \emph{et al.} 2015). This
difficultly is compounded by a few factors, the first being that
sampling ecosystems is not easy. Ecological data is often biased,
noisey, and sparse in both space and time. The current paucity of
ecological data has resulted in much interest in developing global
systems for \emph{ecosystem monitoring} (Makiola \emph{et al.} 2020),
which would systematize the collection of biodiversity data in manner
that makes detecting and predicting change more possible than at the
moment (Urban \emph{et al.} 2021).

\textbf{\emph{TK add mention of interaction prediction here as they
effect dynamics}} The second major challenge in ecological forecasting
is that the underlying dynamics of most ecological processes are unknown
and instead must be inferred from this (sparse) data. Much of the
history of quantitatively modeling ecosystems have been done in the
language of dynamical systems, describing how the value of an observable
state of the system, represented by a vector of numbers
\([x_1, x_2, \dots, x_n]^T = \vec{x}\) changes as over time, yielding
models in the form of differential equations in continuous-time
settings, \(\frac{dx}{dt} = f(x)\), or difference equations in
discrete-time settings, \(x_t = f(x_{t-1})\), where
\(f:\mathbb{R}^n \to \mathbb{R}^n\) is an arbitrary function describing
how the system changes on a moment-to-moment basis (e.g.~in the context
of communities, \(f\) could be Lotka-Voltera, Holling-Type-III or
DeAngelis-Beddington functional response). The form of this functional
response in real systems is effectively unknown, and some forms are
inherently more ``forecastable'' than others (Beckage \emph{et al.}
2011; Chen \emph{et al.} 2019; Pennekamp \emph{et al.} 2019). The
initial success of these forms of models can be traced back to the
larger program of ontological reductionism, which became the default
approach to modeling in the sciences after its early success in physics,
which, by the time ecology was becoming a quantitative science (sometime
in the 20th century, depending on who you ask), became the foundation
for mathematical models in ecology.

@@

We then go on to testing some assumptions of this neutral model with
empirical data. Primarily that we analytically show that our neutral
model, if anything, underestimates the probability of false-negatives if
there are positive associations between species co-occurrence, and we
show these positive associations exist in two sets of spatially
replicated samples of interaction networks (Thompson \& Townsend 2000;
Hadfield \emph{et al.} 2014), fig.~\ref{fig:posassoc}---further I'm
planning to add the field data from the previous chapter into this
analysis once available.

Finally this chapter proposes a simulated annealing method to optimize
the a set of \(n\) points in space to maximize the probability of
detecting an interaction between two species \(a\) and \(b\) with
\emph{known} distributions \(D_a\), \(D_b\).

@@

The second major result is that we analytically show that the this
simulated observation model, by assuming that there is no correlation
between observing two species given that they interaction, actually
underpredicts the realized false-negative interaction rate. We then
demonstrate that this association exists in two empirical systems
fig.~\ref{fig:posassoc}.

@@

At the moment, we have derived phylogenies and SDMs for all the species
present in the Colorado GBIF metaweb. I've also been exploring the data
available from Julian Resasco. The primary constraint on further
progress is that we are waiting on the finalization of a data sharing
agreement with RMBL.

@@

The biggest challenge in implementing simulated annealing in this
context is defining a proposal function for landscape modifications.
This is done by computing the minimum-spanning-tree (MST) of the spatial
nodes, and then proposing corridors that connect nodes that are already
connected in the MST.

@@

\hypertarget{results-3}{%
\subsection{Results}\label{results-3}}

@@

The software as it exists is capable of simulating the biomass dynamics
of arbitrarily large food-webs using Lotka-Volterra, Holling Type-II, or
Holling Type-III functional responses. It currently has methods to
implement Guassian drift, and verious forms of dispersal via
Dispersal.jl. Also occupancy dynamics for Levins metapopulations
(\textbf{levins1967?}), and spatially explicit Hanski metapopulatoins
(\textbf{hanski2001?}). This is most of what needs to exist for the
preceding chapters.

Selection on arbitrary environmental variables in progress, as well as
traits.

>>>>

Within the last several hundred years, human activity has induced rapid
changes in Earth's atmosphere, oceans, and surface. Greenhouse gas
emissions have caused an increase the temperature of both Earth's
terrain and oceans, and both agricultural and urban development has
rapidly reshaped the Earth's land cover. These the bulk of this change
has occurred within the last several hundred years, a geological
instant, inducing a sudden shift in conditions to Earth's climate and
biosphere. As a result \emph{ecological forecasting}---modeling how
ecosystems and their services will change in the future---and then using
these forecasts to make decisions to mitigate the negative consequences
of this change on ecosystems, their functioning, and the services they
provide to humans has emerged as an imperative for ecology and
environmental science (Dietze 2017). However, robust prediction of
ecological processes is, to say the least, quite difficult (Beckage
\emph{et al.} 2011; Petchey \emph{et al.} 2015). This difficultly is
compounded by a few factors, the first being that sampling ecosystems is
not easy. Ecological data is often biased, noisey, and sparse in both
space and time. The current paucity of ecological data has resulted in
much interest in developing global systems for \emph{ecosystem
monitoring} (Makiola \emph{et al.} 2020), which would systematize the
collection of biodiversity data in manner that makes detecting and
predicting change more possible than at the moment (Urban \emph{et al.}
2021).

The second major challenge in ecological forecasting is that the
underlying dynamics of most ecological processes are unknown and instead
must be inferred from this (sparse) data. Much of the history of
quantitatively modeling ecosystems have been done in the language of
dynamical systems, describing how the value of an observable state of
the system, represented by a vector of numbers
\([x_1, x_2, \dots, x_n]^T = \vec{x}\) changes as over time, yielding
models in the form of differential equations in continuous-time
settings, \(\frac{dx}{dt} = f(x)\), or difference equations in
discrete-time settings, \(x_t = f(x_{t-1})\), where
\(f:\mathbb{R}^n \to \mathbb{R}^n\) is an arbitrary function describing
how the system changes on a moment-to-moment basis (e.g.~in the context
of communities, \(f\) could be Lotka-Voltera, Holling-Type-III or
DeAngelis-Beddington functional response). The form of this functional
response in real systems, and whether it is meaningfully non-zero for a
given species interaction, is effectively unknown, and some forms are
inherently more ``forecastable'' than others (Beckage \emph{et al.}
2011; Chen \emph{et al.} 2019; Pennekamp \emph{et al.} 2019). The
initial success of these forms of models can be traced back to the
larger program of ontological reductionism, which became the default
approach to modeling in the sciences after its early success in physics,
which, by the time ecology was becoming a quantitative science (sometime
in the 20th century, depending on who you ask), became the foundation
for mathematical models in ecology.

@@

We then go on to testing some assumptions of this neutral model with
empirical data. Primarily that we analytically show that our neutral
model, if anything, underestimates the probability of false-negatives if
there are positive associations between species co-occurrence, and we
show these positive associations exist in two sets of spatially
replicated samples of interaction networks (Thompson \& Townsend 2000;
Hadfield \emph{et al.} 2014), fig.~\ref{fig:posassoc}---further I'm
planning to add the field data from the previous chapter into this
analysis once available. Finally this chapter proposes a simulated
annealing method to optimize the a set of \(n\) points in space to
maximize the probability of detecting an interaction between two species
\(a\) and \(b\) with \emph{known} distributions \(D_a\), \(D_b\).

@@

The second major result is that we analytically show that the this
simulated observation model, by assuming that there is no association
between observing two species given that they interact, actually under
predicts the realized false-negative interaction rate. We then
demonstrate that this positive association association exists in two
empirical systems fig.~\ref{fig:posassoc}.

@@

At the moment, we have derived phylogenies (fig.~\ref{fig:phylo}) and
SDMs (fig.~\ref{fig:example_sdm}) for all the species present in the
Colorado GBIF metaweb. I've also been exploring the data available from
Julian Resasco. The primary constraint on further progress is that we
are waiting on the finalization of a data sharing agreement with RMBL.

@@

The biggest challenge in implementing simulated annealing in this
context is defining a proposal function for landscape modifications. At
the moment this is done by computing the minimum-spanning-tree (MST) of
the spatial nodes, and then proposing corridors that connect nodes that
are already connected in the MST.

The final component of this chapter is measuring the effect of land-use
change on the robustness of the optimized corridor.

@@

@@

The software as it exists is capable of simulating the biomass dynamics
of arbitrarily large food-webs using Lotka-Volterra, Holling Type-II, or
Holling Type-III functional responses. It currently has methods to
implement Gaussian drift, and verious forms of dispersal via
Dispersal.jl. Also occupancy dynamics for Levins metapopulations
(\textbf{levins1967?}), and spatially explicit Hanski metapopulatoins
(\textbf{hanski2001?}). This is most of what needs to exist for the
preceding chapters. In-progress functionality includes selection (which
affects growth-rate) on arbitrary environmental variables in progress,
as well as traits.
